---
title: "UK Election Polls: Where is the centre?"
format: 
  html:
    theme: spacelab
    self-contained: true
  gfm: 
    output-file: README.md
format-links: false
---

##### *2026-02-21*

---

If you take UK polling data and stack the parties from left to right, who is in the middle parties i.e. which side is capturing more than half of voters?

### A plot of UK general election polls, with parties ordered *roughly* by where they sit on a "right-left" spectrum:

```{r load data, echo = FALSE, message = FALSE}
library(plotly)

# load polling data:
poll_data = read.csv("data/GE_polls_all.csv")

poll_data$date = as.Date(poll_data$date)

party_cols = c(
    "Other_Right",
    "Reform_UKIP", 
    "Conservative", 
    "Lib_Dem", 
    "Labour", 
    "SNP", 
    "Green", 
    "Other_Left"
)


get_weighted_polls <- function(all_poll_data, party_cols, this_sigma = 7) {

    # Dynamically identify party columns (numeric columns except Samplesize)
    # Create the timeline
    all_dates = seq(min(all_poll_data$date), max(all_poll_data$date), by = "day")
    
    # 2. Updated Kernel Calculation
    calc_weighted_avg = function(target_date, all_poll_data, this_sigma, cols) {
        # Time difference in days (Target - Poll Date)
        diffs <- as.numeric(target_date - all_poll_data$date)
        
        # CALCULATE CAUSAL WEIGHTS:
        # If diffs < 0, the poll is in the future relative to target_date.
        # We set those weights to 0.
        time_weights <- exp(-(diffs^2) / (2 * this_sigma^2))
        time_weights[diffs < 0] <- 0 
        
        # Combined weight: Time Decay * Sample Size
        combined_weights <- time_weights * all_poll_data$Samplesize
        
        # Handle the case where no polls exist yet for a date
        if (sum(combined_weights) == 0) {
            return(rep(NA, length(cols)))
        }
        
        # Calculate weighted mean
        res <- sapply(cols, function(p) {
            weighted.mean(all_poll_data[[p]], combined_weights, na.rm = TRUE)
        })
        
        return(res)
    }
    
    calc_weighted_avg(target_date = all_dates[10], all_poll_data, this_sigma, party_cols)
    
    # 3. Apply over the sequence
    weighted_matrix <- t(sapply(all_dates, function(d) {
        calc_weighted_avg(d, all_poll_data, this_sigma, party_cols)
    }))
    
    # 4. Final Formatting
    results <- data.frame(date = all_dates, weighted_matrix)
    names(results)[-1] = party_cols # Ensure party names are preserved
    rownames(results) = NULL
    
    return(results)
}

weighted_polls = get_weighted_polls(poll_data, party_cols = party_cols, this_sigma = 60)

```

```{r stacked_polls, echo = FALSE}
# plot as a stacked area chart

party_cols_named <- setNames(
    c("#ba4cfaff", "#17c8e3ff", "#486cefff", "#ffc400ff", "#e54b4bff", "#fbff16ff", "#4ec94eff", "#f97272ff"),
    party_cols
)



# library(ggplot2)
# ggplot_data = reshape2::melt(weighted_polls, id.vars = "date", measure.vars = party_cols,
#                             variable.name = "Party", value.name = "Percentage")


# #names(ggplot_data)

# ggplot_data$Party = factor(ggplot_data$Party, levels = party_cols)



# ggplot(ggplot_data, aes(x = date, y = Percentage, fill = Party)) +
#     geom_area() +
#     labs(title = "UK Opinion Polls", x = "Date", y
# = "% of respondents") +
#     scale_fill_manual(values = party_cols_named) +
#     # add horizontal line at 0.5
#     geom_hline(yintercept = 0.5, linetype = 1, color = "#111111") +
#     geom_hline(yintercept = 0.45, linetype = 2, color = "#222222") +
#     geom_hline(yintercept = 0.55, linetype = 2, color = "#222222") +
#     geom_hline(yintercept = 0.25, linetype = 2, color = "#555555") +
#     geom_hline(yintercept = 0.75, linetype = 2, color = "#555555") +
#     scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.1), expand = c(0, 0)) +
#     scale_x_date(expand = c(0, 0), date_breaks = "2 years", date_labels = "%Y") +
#     theme_minimal()






# Create plotly figure
fig_stacked = plot_ly()

# Add area traces for each party (in reverse order for proper stacking)
for (i in length(party_cols):1) {
    party = party_cols[i]
    fig_stacked = fig_stacked %>% add_trace(
        x = weighted_polls$date,
        y = weighted_polls[, party],
        type = "scatter",
        mode = "none",
        fill = "tonexty",
        fillcolor = party_cols_named[party],
        #line = list(color = party_cols_named[party]),
        name = party,
        stackgroup = "one",
        groupnorm = ""
    )
}

# Add horizontal reference lines
fig_stacked = fig_stacked %>%
    add_segments(x = min(weighted_polls$date), xend = max(weighted_polls$date),
                y = 0.5, yend = 0.5, line = list(color = "#111111", width = 1),
                showlegend = FALSE, hoverinfo = "skip", inherit = FALSE) %>%
    add_segments(x = min(weighted_polls$date), xend = max(weighted_polls$date),
                y = 0.45, yend = 0.45, line = list(color = "#222222", width = 1, dash = "dash"),
                showlegend = FALSE, hoverinfo = "skip", inherit = FALSE) %>%
    add_segments(x = min(weighted_polls$date), xend = max(weighted_polls$date),
                y = 0.55, yend = 0.55, line = list(color = "#222222", width = 1, dash = "dash"),
                showlegend = FALSE, hoverinfo = "skip", inherit = FALSE) %>%
    add_segments(x = min(weighted_polls$date), xend = max(weighted_polls$date),
                y = 0.25, yend = 0.25, line = list(color = "#555555", width = 1, dash = "dash"),
                showlegend = FALSE, hoverinfo = "skip", inherit = FALSE) %>%
    add_segments(x = min(weighted_polls$date), xend = max(weighted_polls$date),
                y = 0.75, yend = 0.75, line = list(color = "#555555", width = 1, dash = "dash"),
                showlegend = FALSE, hoverinfo = "skip", inherit = FALSE)

# Update layout
fig_stacked = fig_stacked %>% layout(
    title = "UK Opinion Polls",
    xaxis = list(title = "Date", 
                 range = c(min(weighted_polls$date), max(weighted_polls$date)),
                 dtick = "M24"  # tick every 24 months (2 years)
                 ),
    yaxis = list(title = "% of respondents", tickformat = ".0%", 
                 range = c(0, 1), dtick = 0.1),
    hovermode = "x unified"
)

fig_stacked
```

## The standard line chart

Typically polls are presented as a standard line chart like this:

```{r poll_line_chart, echo = FALSE}
# a line plot of the above data
# ggplot(ggplot_data, aes(x = date, y = Percentage, color = Party)) +
#     geom_line() +
#     labs(title = "UK Opinion Polls (2010 - 2026)", x = "Date", y = "% of respondents") +
#     scale_color_manual(values = party_cols_named) +
#     scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.1), expand = c(0, 0)) +
#     scale_x_date(expand = c(0, 0), date_breaks = "2 years", date_labels = "%Y") +
#     theme_minimal()


fig_line = plot_ly()

for (i in seq_along(party_cols)) {
    party = party_cols[i]
    fig_line = fig_line %>% add_trace(
        x = weighted_polls$date,
        y = weighted_polls[, party],
        type = "scatter",
        mode = "lines",
        line = list(color = party_cols_named[party]),
        name = party
    )
}

fig_line = fig_line %>% layout(
    title = "UK Opinion Polls (2010 - 2026)",
    xaxis = list(title = "Date",
                 range = c(min(weighted_polls$date), max(weighted_polls$date)),
                 dtick = "M24"),
    yaxis = list(title = "% of respondents", tickformat = ".0%",
                 #range = c(0, 1), 
                 dtick = 0.1),
    hovermode = "x unified"
)

fig_line
```

Historically with two main parties (Conservative and Labour), just looking at the one with the highest share gives you a good sense of where the public mood is.

But in recent years, parties like Reform and Greens are taking a bigger shares of voting preference, making it hard to see how overall opinion stacks up.

These lines could be misleading. For example a simple narrative you may draw is "Reform have the highest number, therefore they are in the lead".

In a First Past The Post (FPTP) system there is a lot of truth in that in terms of who might win power, but I am always curious whether public opinion in the country as a whole has actually shifted more right, or has the broad left-right balance remained the same?

I think the "stacked" chart gives a slighty clearer view on this, and shows the overall balance is roughly in the same place, but clearly people have shifted to less centrist parties.

## Actual general election results

We can produce the same chart for general election results. These are all the votes in each general election, aggregated across all constituencies. (This is obviously different from a plot of the actual seats won by each party, which would show much bigger shifts due to the FPTP system)

```{r stacked_GE_results, echo = FALSE}
GE_results = read.csv("data/GE_results.csv")

GE_results$date = as.Date(GE_results$date)


GE_party_cols_named <- setNames(
  c("#ba4cfaff", "#17c8e3ff", "#486cefff", "#ffc400ff", "#e54b4bff", "#fbff16ff", "#4ec94eff", "#f97272ff"),
  party_cols
)

# GE_ggplot_data = reshape2::melt(GE_results, id.vars = "date", measure.vars = party_cols,
#                               variable.name = "Party", value.name = "Percentage")


# #names(GE_ggplot_data)

# GE_ggplot_data$Party = factor(GE_ggplot_data$Party, levels = party_cols)



# ggplot(GE_ggplot_data, aes(x = date, y = Percentage, fill = Party)) +
#     geom_area() +
#     labs(title = "UK General Election Results (1918 - 2024)", x = "date", y
#  = "% of voters") +
#     scale_fill_manual(values = GE_party_cols_named) +
#     # add horizontal line at 0.5
#     geom_hline(yintercept = 0.5, linetype = 1, color = "#111111") +
#     geom_hline(yintercept = 0.45, linetype = 2, color = "#222222") +
#     geom_hline(yintercept = 0.55, linetype = 2, color = "#222222") +
#     geom_hline(yintercept = 0.25, linetype = 2, color = "#555555") +
#     geom_hline(yintercept = 0.75, linetype = 2, color = "#555555") +
#     scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.1), expand = c(0, 0)) +
#     scale_x_date(expand = c(0, 0)) + #, date_breaks = "2 years", date_labels = "%Y") +
#     theme_minimal()




fig_GE_stacked = plot_ly()

for (i in length(party_cols):1) {
    party = party_cols[i]
    fig_GE_stacked = fig_GE_stacked %>% add_trace(
        x = GE_results$date,
        y = GE_results[, party],
        type = "scatter",
        mode = "none",
        fill = "tonexty",
        fillcolor = GE_party_cols_named[party],
        name = party,
        stackgroup = "one",
        groupnorm = ""
    )
}

fig_GE_stacked = fig_GE_stacked %>%
    add_segments(x = min(GE_results$date), xend = max(GE_results$date),
                 y = 0.5, yend = 0.5, line = list(color = "#111111", width = 1),
                 showlegend = FALSE, hoverinfo = "skip", inherit = FALSE) %>%
    add_segments(x = min(GE_results$date), xend = max(GE_results$date),
                 y = 0.45, yend = 0.45, line = list(color = "#222222", width = 1, dash = "dash"),
                 showlegend = FALSE, hoverinfo = "skip", inherit = FALSE) %>%
    add_segments(x = min(GE_results$date), xend = max(GE_results$date),
                 y = 0.55, yend = 0.55, line = list(color = "#222222", width = 1, dash = "dash"),
                 showlegend = FALSE, hoverinfo = "skip", inherit = FALSE) %>%
    add_segments(x = min(GE_results$date), xend = max(GE_results$date),
                 y = 0.25, yend = 0.25, line = list(color = "#555555", width = 1, dash = "dash"),
                 showlegend = FALSE, hoverinfo = "skip", inherit = FALSE) %>%
    add_segments(x = min(GE_results$date), xend = max(GE_results$date),
                 y = 0.75, yend = 0.75, line = list(color = "#555555", width = 1, dash = "dash"),
                 showlegend = FALSE, hoverinfo = "skip", inherit = FALSE)

fig_GE_stacked = fig_GE_stacked %>% layout(
    title = "UK General Election Results (1918 - 2024)",
    xaxis = list(title = "Date",
                 range = c(min(GE_results$date), max(GE_results$date)),
                 dtick = "M120"  # tick every 10 years
                 ),
    yaxis = list(title = "% of voters", tickformat = ".0%",
                 range = c(0, 1), dtick = 0.1),
    hovermode = "x unified"
)

fig_GE_stacked

```

Here again we see that the centre is fairly stable, and winning party rarely makes it over the 50% line.


## Thoughts/Observations

#### Shifting values and the political marketplace

The stablilty at first was surprising to me, I was expecting to see bigger swings.

But maybe this is because the parties are shifting their views to match public opinion (and vice versa). A party that is misaligned with its target voters, would lose share of opinion over time. The competing forces of chasing votes and the parties ideology leads to a stable balance of left vs right.  

The underlying views and ideas held by the parties and the public are of course shifting over time (opinions on gay rights and immigration have shifted a lot over the years). But you wouldn't necessarily see these kinds of idealogical shifts in this data if parties move in line with public opinion.

How those idealogical views are changing would require tracking these specific topics. The Polical Compass website has tried to map this shift in opinions over time (see [https://politicalcompass.org/](https://www.politicalcompass.org/uk2017)).

And Yougov have also done some reseach trying to track these kinds of shifts. Here are a few examples:

-   [Yougov: Where do Britons see politicians, parties and themselves on the left-right spectrum](https://yougov.co.uk/politics/articles/52080-where-do-britons-see-politicians-parties-and-themselves-on-the-left-right-spectrum)

-   [Yougov: Left-wing vs right-wing: It's complicated](https://yougov.co.uk/politics/articles/24767-left-wing-vs-right-wing-its-complicated)

-   [Yougov: UKIP voters put themselves left of Tories](https://yougov.co.uk/politics/articles/11200-ukip-voters-put-themselves-left-tories)

#### Should most elections have been won by a left wing/progressive party?

If you consider the Lib dems as a left of centre/progressive party, then looking at the stacked charts, and focusing on that central 50% line, then you might say that most UK elections should have been won by a left wing/progressive party, and indeed that may have happened under proportional representation.

A counter arguement would again be the idea that parties shift to match voters. Hence if we had PR (or the lib dems just didn't exist) and the conservatives kept loosing elections, then they would presumably shift their position more to the left and a new equilibrium would be found.

Either way, you could argue that historically the left wing vote has been split, leading to those views being underrepresented in parliament. The rise in Reform may potentially split the right in a similar way.

#### Undecided voters

I would love to make a version of this that includes undecided voters. Again, this wouldn't tell you who is going to win, but it would just be interesting to see where the weight of public opinion lies.


## Notes

#### Merged parties

I've merged some parties for simplicity. In particular I've combined the data of UKIP, Brexit and Reform, and for the general election data combined the historic libral party with the modern Lib Dems.

#### Other Left and Other Right

For smaller parties not explicitly named, I've tried to put them as "other left" or "other right", but where unknown I've split the share between these two groups. The share is so low for these groups it doesn't noticiably effect the centre position.

Over time different parties are reported by polling companies. For example SNP has been reported since 2015, but before then wasn't widely reported. So before that point, they are counted in the "Other Left" category, hence the discontinuity on the chart.


#### Party's position on the spectrum

The assignment of parties on the left right spectrum could be argued (is SNP left or right of Labour?), but I really just wanted to roughly put them at one side of the fence or the other.

#### Excluding Northern Ireland, sorry!

All of the above excludes Northern Ireland, as NI parties are generally excluded from the main GE polls.

## References

All the polling data has come via wikipedia:

<https://en.wikipedia.org/wiki/Opinion_polling_for_the_2015_United_Kingdom_general_election>

<https://en.wikipedia.org/wiki/Opinion_polling_for_the_2017_United_Kingdom_general_election>

<https://en.wikipedia.org/wiki/Opinion_polling_for_the_2019_United_Kingdom_general_election>

<https://en.wikipedia.org/wiki/Opinion_polling_for_the_2024_United_Kingdom_general_election>

<https://en.wikipedia.org/wiki/Opinion_polling_for_the_next_United_Kingdom_general_election>

The general election data has come the commons library:

<https://commonslibrary.parliament.uk/research-briefings/cbp-8647/>

##### *Keith Simpson, 2026*